{\large Master Thesis}\\{\normalsize on

### 2011-07-19 doc/report.tex
m

\title{Algorithms for Longest Common Extensions}

\author{Jesper Kristensen, s062397}

@@

\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers of $\sigma$ distinct values is written as $\sortt$.

@@

For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path. If a node $u$ has height $h$, $u$'s ladder contains at least the last $h$ ancestors of $u$. The combined size of the ladders is at most $2n$.~\cite{level-ancestor}

@@

To perform a LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
\begin{samepage}
\begin{enumerate}
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless and $\ell = k-1$.
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
\item When $H_\ell[i+v] \neq H_\ell[j+v]$, decrement $\ell$ by one and repeat step 2.
\item Stop when $\ell = 0$ and $H_\ell[i+v] \neq H_\ell[j+v]$.
\end{enumerate}

The LCE value is now $v = \LCE(i,j)$. An example is shown in \pref{fig:fingerprint-query}.
\end{samepage}

@@

For larger values of $k$, we can attempt to optimize memory access accross different levels. When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$. Neither of these positions are next to $H_\ell[i]$ in memory. We can optimize for the case where we reach $H_{\ell-1}[i]$ by flipping the direction in which we store the fingerprints, such that all fingerprints $H_\ell[i]$ for all $\ell$ and a specific $i$ are stored next to each other. This will not improve the asymptotic I/O bound, but could improve practical query times.

@@

Sometimes it can be useful to store a string in a compressed format and query the compressed string directly without using extra space and time to first decompress the string. In this section we explore the LCE problem on strings compressed using LZ78 compression. We look at solutions, which require space proportional to the compressed size of a string, and are able to answer LCE queries fast.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LZ78 definition\label{sec:lz-definition}}
A string $s$ of length $N$ has a corresponding \textit{compressed string} $z$ of compressed length $n$, where $z$ is a list of \textit{compression elements}, $z = z_1, z_2, ..., z_n$. Each compression element $z_x$ is a pair $(r_x, \alpha_x)$, where the reference, $\zreference_z(x)=r_x$, is an index of an earlier compression element in the compressed string or zero, i.e. $0 \leq \zreference_z(x) < x$, and the label, $\zlabel_z(x)=\alpha_z$, is a character from the original uncompressed string $s$. The \zphrase\ of a compression element is defined as:
\begin{align*}
\zphrase_z(x) =
\begin{cases}
    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference \neq 0\\
    \zlabel_z(x) & \textrm{otherwise}
\end{cases}
\end{align*}
where "$\cdot$" denotes string concatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.

We can easily get the decompressed string $s$ from the compressed string $z$ using the definitions. To get the compressed string $z$ from the original string $s$ we can look at the compression elements as a tree, where the reference of each compression element corresponds to the parent pointer of each node in the tree. We consume $s$ from left to right while we build the tree:
\begin{enumerate}
\item Let the tree be single special root node $z_0$, let $i = 1$ and $c=z_0$.
\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create the child node $u$, set $\zlabel(u)=s[i]$ and set $c=z_0$.
\item Increment $i$ by one and repeat from step 2 until the entire string is consumed when $i = n$.
\end{enumerate}
The compressed string $z$ is now the tree except the special root node $z_0$.

@@

\begin{proof}
To find the character $s[i]$, we can first use a Predecessor query to find out which compression element $s[i]$ is in. We can then use a Level Ancestor query to find the correct character from the found compression element.

@@

\begin{proof}
Whenever \proc{DirectComp} reads a character of the string $s$, we instead use the random access described in \pref{lem:lz-random-access} to read the character from the compressed string. The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$. In subsequent iterations of \proc{DirectComp} we know that we always need the same compression element or the following compression element as we used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the compression element. The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
\end{proof}

@@

A simpler way of doing LCE on compressed a string is to reverse the string and then run \proc{DirectComp} in the reverse direction on the reversed string, which gives the same result since reversing a string twice gives the original string. Where \proc{DirectComp} compares $s[i+v]$ against $s[j+v]$ in each iteration of its loop, reversed \proc{DirectComp} compares $s[N+1-i-v]$ against $s[N+1-j-v]$ in each iteration. This is simpler than using \proc{DirectComp} directly on the compressed string, because Level Ancestor queries are no longer needed at each iteration, since we can use the reference of the current compression element as follows, where the position of each of $i$ and $j$ is stored as a pair $(x,y)$:
\begin{align*}
\proc{PreviousPosition}(x, y)=&
\begin{cases}
(\zreference(x), y) & \textrm{if} ~ \zreference(x) \neq 0 \\
(y-1, y-1) & \textrm{if} ~ y > 1\\
\textit{End of string} & \textrm{otherwise}
\end{cases}\\
\proc{Character}(x, y)=&~\zlabel(x)\\
\proc{Position}(i)=&~(x,y)~\textrm{where}\\
&~~y=\proc{Predecessor}(i)~\textrm{and}\\
&~~x=\proc{Level-Ancestor}(i-\proc{Start-Of}(y), y)
\end{align*}
We then only need one Predecessor and one Level Ancestor query to initialize each of the two pairs.

@@

Reversing the string can be done in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively. To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-definition} reads its input string one character at a time from left to right and uses $O(n)$ space for its tree of compression elements.

@@

We have found one case where we can make compressed \proc{DirectComp} faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time. We explain this in terms of the reversed compressed \proc{DirectComp}, as it is simpler, but it also works without the string reversal. Reversed compressed \proc{DirectComp} starts with two positions $(x_i,y_i)$ and $(x_j,y_j)$ corresponding to the two positions $i$ and $j$ in $\LCE(i,j)$. At each iteration, it then updates $(x_i,y_i)$ and $(x_j,y_j)$ to find the previous two positions. If $x_i = x_j$ we know that we can add $d(x_i)$ to the LCE value, where $d(x_i)$ is the depth of node $x_i$ in the tree of compression elements. Compressed \proc{DirectComp} can then compare $d(x_i)$ characters in constant time.

@@

We can look at the other LCE algorithms we have examined in addition to \proc{DirectComp}, and see if we can apply them to LZ-compressed strings. We have not found a way to apply \proc{SuffixNca} or \proc{LcpRmq}, as the string we want to find LCE values on still has N suffixes, and we have not found a way to relate the compression rate of the NCA structure or suffix array to the compression rate of the string.

Fingerprinting does not compress well either. We could LC-compress each level of fingerprints just like the original string is compressed, but as the substrings fingerprinted becomes longer at each elvel, the compression ration becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.

@@

We have looked into the cache performance of compressed \proc{DirectComp}. If the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case. When the string is LZ-compressable, the I/O pattern becomes much worse. If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be next to each other in memory. This can happen many times, which gives $O(N)$ I/O's in worst case.

@@

\paragraph{Paths}
A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on paths:
\begin{itemize}
\item The path $p=(u,w)$ has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
\item The node $p[x]$ is a node in $p=(u,w)$, such that the distance between $u$ and $p[x]$ is $x$. The start and end nodes of the path are $p[0]=u$ and $p[e]=w$. We can calculate $p[x]$ in constant time as $p[x]=\LA(d(u)+x,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
\item The subpath $p[x\twodots y]=(p[x],p[y])$ of a path $p$ is the path, which starts at node $p[x]$ and ends at node $p[y]$. The entire path is $p = p[0,e]$.
\item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
\end{itemize}

@@

We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\zlabel(p_i[1])$ to $\zlabel(p_j[1])$, then compare $\zlabel(p_i[2])$ to $\zlabel(p_j[2])$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.

@@

We can adapt our \fprintk\ algorithm on strings to work on trees, which we call \tfprintk. If we store the fingerprint of $s[i\twodots i+t_\ell]$ in $H_\ell[i+t_\ell]$ instead of $H_\ell[i]$, then the \fprintk\ query directly adapts to \tfprintk. If we wanted to use the same preprocessing as we use for \fprintk\ on strings, we would have to adapt the suffix array to apply to trees, and find a way to construct it in $\sortt$ time, which we have not looked at. Instead we can preprocess the fingerprints of the tree with Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}.

@@

\fixme{Write out the query in full}

\begin{lemma}
The \tfprintk\ query finds the correct tree LCE value in $O(k\cdot n^{1/k})$ time.
\end{lemma}
\begin{proof}
We have changed nothing from the \fprintk\ query on strings, which negatively affects the query time, so the query time is the same.

% The combination of Level Ancestor and the definition that a fingerprint relates to a substring from a given node to an ancestor that node, ensures that the fingerprints we choose to compare are the fingerprints, which represents the strings on the paths from $u_i$ to $w_i$ and from $u_j$ to $w_j$.

\fixme{Improve this paragraph}
At any given iteration at level $\ell$, we know that $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and we now check if $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$. From our definition of fingerprints in a tree, these two labels have fingerprints $H_\ell[p_i[v+t_\ell]]$ and $H_\ell[p_j[v+t_\ell]]$, which is what our LCE query uses.
\end{proof}

@@

We use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to preprocess the \tfprintk\ data structure. When $t_\ell$ is $t_{\ell-1} < t_\ell \leq 2t_{\ell-1}$, we preprocess level $\ell$ like this:
\begin{enumerate}
\item Make pairs of fingerprints $H_\ell[u] = (H_{\ell-1}[LA(d(u)-t_\ell+t_{\ell-1},u)], H_{\ell-1}[u])$ for all $u$ where $d(u) \geq t_\ell$.
\item Sort the pairs lexicographically.
\item Replace each pair with a new fingerprint, by going through them in lexicographical order, while updating the fingerprint whenever the pair changes.
\end{enumerate}

@@

The \proc{HeavyPathLce} data structure consists of the following:
\begin{itemize}
\item An input tree $T$ with a heavy path decomposition.
\item A string $s$, which is a concatenation of strings of all light edges of $T$, where the string of a light edge is the label of the light edge concatenated with the label of the heavy path, which follows that light edge. \pref{fig:tree-heavy-ds} shows an example of $s$.
\item For a node $u$, $\zend(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ is the depth of $u$, and $\zindex(u)$ is the index in $s$, which stores $\zlabel(u)$.
\item Data structures for constant time NCA on $T$, constant time LA on $T$, and constant time LCE on $s$.
\end{itemize}

@@

\begin{lemma}
We can preprocess the \proc{HeavyPathLce} data structure in $O(\sortt)$ time.
\end{lemma}
\begin{proof}
Preprocessing LCE on $s$ using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time. If we walk through the tree and for each leaf append the light edge string ending in that leaf to $s$, we can construct $s$ in $O(n)$ time. The rest of the data structures can be preprocessed in $O(n)$ time.
\end{proof}

@@

\begin{theorem}
\proc{TreeLcpRmq} and \proc{TreeSuffixNca} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(1)$ query time and $O(n^2)$ space.
\end{theorem}
\begin{proof}
\proc{LcpRmq} relies on accessing two suffixes in SA, $i$ and $j$. In \proc{TreeLcpRmq} we change the two suffixes to paths $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways. We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the original LCE query in $O(1)$ time, but even with this modification, a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each path. The SA and LCP arrays would therefore be of length $O(n^2)$. When we look at \proc{TreeSuffixNca}, the same argument holds for the necessary number of leafs in the suffix tree.
\end{proof}

@@

\proc{LadderLce}, whic we describe in the next section, finds the LCE value of the first half of $p_i$ and $p_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm. We have not found a constant time solution to find the LCE value of the second half of these paths, in the same way jump pointers is used to jump from $u$ node to the half way node node in the Level Ancestor algorithm. We have thus not found a constant time linear space solution for LCE on trees.

@@

The \proc{LadderLce} data structure consists of the following:
\begin{itemize}
\item An input tree $T$ with a ladder decomposition.
\item A strings $s$, which is a concatenation of ladder labels of each ladder in $T$, where the label of a ladder is the concatenation of labels of each node in the ladder, ordered ascending by the depth of the nodes. \pref{fig:ladder-ds} shows an example of $s$.
\item For a node $u$, $\zindex(u)$ is an index in $s$ within the ladder label of $u$'s ladder, where $\zlabel(u)$ is stored.
\item Data structures for constant time LA on $T$ and constant time LCE on $s$.
\end{itemize}

@@

The \proc{LadderLce} query starts with $v=0$ and repeats the following:
\begin{enumerate}
\item Stop when $v = e_i$ or $v = e_j$ and return $v$.
\item Find $m_i = \lceil\zlength(p_i[v\twodots e_i])/2\rceil$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
\fixme{afrunding}
\item Find $m_j = \lceil\zlength(p_j[v\twodots e_j])/2\rceil$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
\item Find $v' = \min\{LCE_s(i,j),m_i,m_j\}$
\item Increment $v$ by $v'$.
\item Stop if $v' < \min\{m_i,m_j\}$ and return $v$.
\end{enumerate}

\begin{lemma}
The \proc{LadderLce} query finds the correct LCE value.
\end{lemma}
\begin{proof}
In step two, $i$ refers to a position in $p_i[v+m_i]$'s ladder, because the height of $p_i[v+m_i]$ is at least $m_i-1$. Therefore $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$, and similarly $s[j\twodots j+m_i-1] = \zlabel(p_j[v\twodots v+m_j])$.

By the same argument as we use for \proc{HeavyPathLce}, the algorithm maintains the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and gives the correct LCE value.
\end{proof}

>>>>

\title{{\large Master Thesis}\\{\normalsize on}\\Algorithms for Longest Common Extensions}

\author{Jesper Kristensen\\s062397\\DTU Informatics\\Technical University of Denmark}

\date{August 1, 2011}

@@

\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers within an universe of $\sigma$ is written as $\sortt$.

@@

For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path. If a node $u$ has height $h$, $u$'s ladder contains all the ancestors of $u$ whose distance to $u$ is at most $h+1$. The combined size of the ladders is at most $2n$.~\cite{level-ancestor}

@@

To perform a LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
\begin{samepage}
\begin{enumerate}
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless and $\ell = k-1$.
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
\item Stop and return $v$ when $\ell = 0$, otherwise decrement $\ell$ by one and go to step two.
\end{enumerate}

An example of a query is shown in \pref{fig:fingerprint-query}.
\end{samepage}

@@

For larger values of $k$, we can attempt to optimize memory access across different levels. When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$. Neither of these positions are next to $H_\ell[i]$ in memory. We can optimize for the case where we reach $H_{\ell-1}[i]$ by flipping the direction in which we store the fingerprints, such that all fingerprints $H_\ell[i]$ for all $\ell$ and a specific $i$ are stored next to each other. This will not improve the asymptotic I/O bound, but could improve practical query times.

@@

Sometimes it can be useful to store a string in a compressed format and query the compressed string directly without using extra space and time to first decompress the string. In this section we explore the LCE problem on strings compressed using LZ78 compression. We look at solutions, which require space relative to the compressed size of a string, and we try to find solutions, which can answer LCE queries fast.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LZ78 definition}
A string $s$ of length $N$ has a corresponding \textit{compressed string} $z$ of compressed length $n$, where $z$ is a list of \textit{compression elements}, $z = z_1, z_2, ..., z_n$. Each compression element $z_x$ is a pair $(r_x, \alpha_x)$, where the reference, $\zreference_z(x)=r_x$, is an index of an earlier compression element in the compressed string or zero, i.e. $0 \leq \zreference_z(x) < x$, and the label, $\zlabel_z(x)=\alpha_z$, is a character from the original uncompressed string $s$. The \zphrase\ of a compression element is defined as:
\begin{align*}
\zphrase_z(x) =
\begin{cases}
    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference_z(x) \neq 0\\
    \zlabel_z(x) & \textrm{otherwise}
\end{cases}
\end{align*}
where "$\cdot$" denotes string concatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.

\subsubsection{Compressing and Decompressing\label{sec:lz-comp-alg}}

We can easily get the decompressed string $s$ from the compressed string $z$ using the definitions. To get the compressed string $z$ from the original string $s$ we can look at the compression elements as a tree, where the reference of each compression element corresponds to the parent pointer of each node in the tree. We consume $s$ from left to right while we build the tree:
\begin{enumerate}
\item Let the tree be a single special root node $z_0$, let $i = 1$ and $c=z_0$.
\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create a new node $u$ as a child of $c$, set $\zlabel(u)=s[i]$, and set $c=z_0$.
\item Increment $i$ by one and repeat from step two until the entire string is consumed when $i = N$.
\end{enumerate}
The compressed string $z$ is now the tree except the special root node $z_0$.

@@

\begin{proof}
To find the character $s[i]$, we can first use a Predecessor query to find out which phrase $s[i]$ is in. We can then use a Level Ancestor query to find the correct character from the found phrase.

@@

\begin{proof}
Whenever \proc{DirectComp} reads a character of the string $s$, we instead use the random access described in \pref{lem:lz-random-access} to read the character from the compressed string. The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$. In subsequent iterations of \proc{DirectComp} we always need the same phrase or the phrase following the one we used used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the phrase. The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
\end{proof}

@@

A simpler way of doing LCE on a compressed string is to reverse the string and then run \proc{DirectComp} in the reverse direction on the reversed string, which gives the same result since reversing a string twice gives the original string. Where \proc{DirectComp} compares $s[i+v]$ against $s[j+v]$ in each iteration of its loop, reversed \proc{DirectComp} compares $s^R[N+1-i-v]$ against $s^R[N+1-j-v]$ in each iteration. This is simpler than using \proc{DirectComp} directly on the compressed string, because Level Ancestor queries are no longer needed at each iteration, since we can use the reference of the current compression element as follows, where the position of each of $i$ and $j$ is stored as a pair $(x,y)$:
\begin{align*}
\proc{PreviousPosition}(x, y)=&
\begin{cases}
(\zreference(x), y) & \textrm{if} ~ \zreference(x) \neq 0 \\
(y-1, y-1) & \textrm{if} ~ y > 1\\
\textit{End of string} & \textrm{otherwise}
\end{cases}\\
\proc{Character}(x, y)=&~\zlabel(x)\\
\proc{Position}(i)=&~(x,y)~\textrm{where}\\
&~~y=\proc{Predecessor}(i)~\textrm{and}\\
&~~x=\proc{Level-Ancestor}(i-\proc{Start-Of}(y), y)
\end{align*}
We then only need one Predecessor and one Level Ancestor query to initialize each of the two pairs.

@@

We can reverse the string in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively. To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-comp-alg} reads its input string one character at a time from left to right and uses $O(n)$ space for its tree of compression elements.

@@

We have found one case where we can make the reversed compressed \proc{DirectComp} faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time. Reversed compressed \proc{DirectComp} starts with two positions $(x_i,y_i)$ and $(x_j,y_j)$ corresponding to the two positions $i$ and $j$ in $\LCE(i,j)$. At each iteration, it then updates $(x_i,y_i)$ and $(x_j,y_j)$ to find the previous two positions. If $x_i = x_j$ we know that we can add $d(x_i)$ to the LCE value, where $d(x_i)$ is the depth of node $x_i$ in the tree of compression elements. Reversed compressed \proc{DirectComp} can then compare $d(x_i)$ characters in constant time.

@@

We can look at the other LCE algorithms we have examined in addition to \proc{DirectComp}, to see if we can apply them to LZ-compressed strings. We have not found a way to apply \proc{SuffixNca} or \proc{LcpRmq}, as the string we want to find LCE values on still has N suffixes, and we have not found a way to relate the compression rate of the NCA structure or suffix array to the compression rate of the string.

Fingerprinting does not compress well either. We could LZ-compress each level of fingerprints just like the original string is compressed, but as the substrings fingerprinted becomes longer at each level, the compression ratio becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.

@@

We have looked into the cache performance of compressed \proc{DirectComp}. If we ignore the predecessor and level ancestor queries, and the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, compressed \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case. When the string has a better LZ-compression ratio, the I/O pattern becomes much worse. If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be next to each other in memory. This can happen many times, which gives $O(N)$ I/O's in worst case.

@@

\paragraph{Paths}
A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on paths:
\begin{itemize}
\item The path $p=(u,w)$ has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
\item The node $p[x]$ is a node in $p=(u,w)$, such that the distance between $u$ and $p[x]$ is $x$. The start and end nodes of the path are $p[0]=u$ and $p[e]=w$. We can calculate $p[x]$ in constant time as $p[x]=\LA(d(u)+x,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
\item The subpath $p[x\twodots y]=(p[x],p[y])$ of a path $p$ is the path, which starts at node $p[x]$ and ends at node $p[y]$. The entire path is $p = p[0\twodots e]$.
\item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
\end{itemize}

@@

We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\zlabel(p_i[1])$ to $\zlabel(p_j[1])$, then compare $\zlabel(p_i[2])$ to $\zlabel(p_j[2])$ and so on until the two characters differ or either $p_i[e_i]$ or $p_j[e_j]$ is reached.

@@

We can adapt our \fprintk\ algorithm on strings to work on trees, which we call \tfprintk. If we store the fingerprint of $s[i\twodots i+t_\ell-1]$ in $H_\ell[i+t_\ell]$ instead of $H_\ell[i]$, then the \fprintk\ query directly adapts to \tfprintk. If we wanted to use the same preprocessing as we use for \fprintk\ on strings, we would have to adapt the suffix array to apply to trees, and find a way to construct it in $\sortt$ time, which we have not looked at. Instead we can preprocess the fingerprints of the tree with Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}.

@@

In the \tfprintk\ query, start with $v=0$ and $\ell=0$, then do the following steps:
\begin{samepage}
\begin{enumerate}
\item As long as $v+t_\ell\leq e_i\lor v+t_\ell\leq e_j$ and $H_\ell[p_i[v+t_\ell]] = H_\ell[p_j[v+t_\ell]]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless $\ell = k-1$.
\item As long as  $v+t_\ell\leq e_i\lor v+t_\ell\leq e_j$ and $H_\ell[p_i[v+t_\ell]] = H_\ell[p_j[v+t_\ell]]$, increment $v$ by $t_\ell$ and repeat this step.
\item Stop and return $v$ when $\ell = 0$, otherwise decrement $\ell$ by one and go to step two.
\end{enumerate}
\end{samepage}

\begin{lemma}
The \tfprintk\ query finds the correct tree LCE value.
\end{lemma}
\begin{proof}
The query algorithm keeps the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$. At any given step at level $\ell$, we check if $H_\ell[p_i[v+t_\ell]]=H_\ell[p_j[v+t_\ell]]$, which by the definition of our data structure is equivalent to $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$. If the equality holds, we know that $\zlabel(p_i[0\twodots v+t_\ell])=\zlabel(p_j[0\twodots v+t_\ell])$, and we maintain our invariant when we increment $v$ by $t_\ell$.

We stop when $v+1>e_i$, $v+1>e_j$, or $H_0[p_i[v+1]] \neq H_0[p_j[v+1]]$, at which point $v=\LCE(p_i,p_j)$.
\end{proof}

\begin{lemma}
The \tfprintk\ query takes $O(k\cdot n^{1/k})$ time.
\end{lemma}
\begin{proof}
We have changed nothing from the \fprintk\ query on strings, which negatively affects the query time, so the query time is the same.
\end{proof}

@@

We use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to preprocess the \tfprintk\ data structure. When $t_\ell$ is $t_{\ell-1} < t_\ell \leq 2t_{\ell-1}$, we preprocess level $\ell$ like this:
\begin{enumerate}
\item Make pairs of fingerprints $H_\ell[u] = (H_{\ell-1}[LA(d(u)-t_\ell+t_{\ell-1},u)], H_{\ell-1}[u])$ for all nodes $u$ in $T$, where $d(u) \geq t_\ell$.
\item Sort the pairs lexicographically.
\item Replace each pair with a new fingerprint, by going through them in lexicographical order, while updating the fingerprint whenever the pair changes.
\end{enumerate}

@@

The \proc{HeavyPathLce} data structure consists of the following:
\begin{itemize}
\item An input tree $T$ with a heavy path decomposition.
\item A string $s$, which is a concatenation of strings of all heavy paths of $T$, where the string of a heavy path is the label of the light edge leading to the heavy path concatenated with the label of the heavy path. \pref{fig:tree-heavy-ds} shows an example of $s$.
\item For a node $u$, $\zend(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ is the depth of $u$, and $\zindex(u)$ is the index in $s$, which stores $\zlabel(u)$.
\item Data structures for constant time NCA on $T$, constant time LA on $T$, and constant time LCE on $s$.
\end{itemize}

@@

\begin{lemma}
We can preprocess the \proc{HeavyPathLce} data structure in $O(\sortt)$ time.
\end{lemma}
\begin{proof}
Preprocessing LCE on $s$ using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time. If we walk through the tree and for each leaf append the heavy path string ending in that leaf to $s$, we can construct $s$ in $O(n)$ time. The rest of the data structures can be preprocessed in $O(n)$ time.
\end{proof}

@@

\begin{theorem}
\proc{TreeLcpRmq} and \proc{TreeSuffixNca} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(1)$ query time and $O(n^2)$ space.
\end{theorem}
\begin{proof}
\proc{LcpRmq} relies on accessing two suffixes in SA, $i$ and $j$. In \proc{TreeLcpRmq} we change the two suffixes to paths $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways. We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the original LCE query in $O(1)$ time, but even with this modification, a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each path. The SA and LCP arrays in \proc{TreeLcpRmq} would therefore be of length $O(n^2)$. When we look at \proc{TreeSuffixNca}, the same argument holds for the necessary number of leafs in the suffix tree.
\end{proof}

@@

\proc{LadderLce}, which we describe in the next section, finds the LCE value of the first half of $p_i$ and $p_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm. We have not found a constant time solution to find the LCE value of the second half of these paths, in the same way jump pointers is used to jump from node $u$ to the half way node in the Level Ancestor algorithm. We have thus not found a constant time linear space solution for LCE on trees.

@@

The \proc{LadderLce} data structure consists of the following:
\begin{itemize}
\item An input tree $T$ with a ladder decomposition.
\item A strings $s$, which is a concatenation of ladder labels of each ladder in $T$, where the label of a ladder is the concatenation of labels on each edge leading to a node in the ladder, ordered ascending by the depth of the nodes. \pref{fig:ladder-ds} shows an example of $s$.
\item For a node $u$, $\zindex(u)$ is an index in $s$ within the ladder label of $u$'s ladder, where $\zlabel(u)$ is stored.
\item Data structures for constant time LA on $T$ and constant time LCE on $s$.
\end{itemize}

@@

The \proc{LadderLce} query starts with $v=0$ and repeats the following:
\begin{enumerate}
\item Stop when $v = e_i$ or $v = e_j$ and return $v$.
\item Find $m_i = \lceil\zlength(p_i[v\twodots e_i])/2\rceil$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
\item Find $m_j = \lceil\zlength(p_j[v\twodots e_j])/2\rceil$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
\item Find $v' = \min\{LCE_s(i,j),m_i,m_j\}$
\item Increment $v$ by $v'$.
\item Stop if $v' < \min\{m_i,m_j\}$ and return $v$.
\end{enumerate}

\begin{lemma}
The \proc{LadderLce} query finds the correct LCE value.
\end{lemma}
\begin{proof}
In step two, $\zlength(p_i[v\twodots v+m_i]) +1 \leq \zlength(p_i[v+m_i\twodots e_i])\leq h(p_i[v+m_i])$, where $h$ is the height of a node. For a node $u$ of height $h$, the ladder decomposition guarantees that $u$'s ancestor with distance $h+1$ to $u$ is in $u$'s ladder. Therefore, $p_i[v+1]$ is in $p_i[v+m_i]$'s ladder, and $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$. Similarly in step three, $s[j\twodots j+m_i-1] = \zlabel(p_j[v\twodots v+m_j])$.

By the same argument we use for \proc{HeavyPathLce}, the algorithm maintains the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and gives the correct LCE value.
\end{proof}
